Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-11-19 article/article.tex
adiantada escrita e scripts de analise

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z \textipa{\!R}].

Syllables are of the form (C)V(N):
an optional consonant, a vowel and an optional nasal consonant.
Non word-initial syllables must follow the pattern CV(N).
The following sequences are forbidden: ji, wu, wo, ti, mn, nm, mm,
nn.

@@

\subsection{Statistics of the vocabulary}
In~\cite{pije} there is statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

As described in Section~\cite{basics},
there is only 14 letters,
and phonemes respect a few rules.
There are 120 different words in the official vocabulary,
4 of them having synonyms.
A total of 124 tokens.
Not counting proper nouns (names)
and punctuation.
They include X nouns, Y verbs, Z adjectives,
W prepositions, Y particles, K pre-verbs.
Of which one might distinguish only between
the particles and the other J words that
might be used anywhere a particle will not.

Most often vowels,
consonants, consonant vs vowel.
Most often phonemes in general an specific positions..
Possible phonemes given the rules.
Comparison of the occurring against the possible.

Syntactic structures:
Possibilities on noun, verb and prepositional phrases
and in sentences.
Counting along the number of words.

\subsection{Synthesis of text}
number of elisoes

@@

>>>>

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z \textipa{\!R}].
Especially for poetry, one might consider
j and w to be compound vowels
(e.g. j is 'i' and w is 'u'),
or as consonants.

Syllables are of the form (C)V(N):
an optional consonant, a vowel and an optional nasal consonant.
Non word-initial syllables must follow the pattern CV(N).
The following sequences are forbidden: ji, wu, wo, ti, nm, nn.

@@

\subsection{Statistics of the vocabulary}
In~\cite{pije} are statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

As described in Section~\cite{basics},
there is only 14 letters,
and phonemes respect a few rules.
There are 120 different words in the official vocabulary,
4 of them having synonyms.
A total of 124 tokens.
Not counting proper nouns (names)
and punctuation.
Table~\ref{tab:pos} shows the number of words
related to each POS tag specified in the dictionary.

\input{pos}

Vowel frequencies are as shown in Table~\ref{tab:vowels}
for starting, end and internal position.
Consonants are shown in Table~\ref{tab:cons}
for meaningful values.
Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
The high incidences of 'l' and 'a' favor
the naturalness of the language prosody
as it resembles lallation.

Most often phonemes are in
Table~\ref{tab:phon}.
A toolbox for obtaining such statistics and sets
of words is in~\cite{tokipona}.

Given the rules given in Section~\ref{phonology}
(of 15 letters, 5 vowels, (C)V(N) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
X words are possible with 1 syllable,
Y with 2 syllables, Z with 3 syllables
and W with 4 syllables.
These are shown in Table~\ref{tab:pw}
with fraction of the space actually used in Toki Pona.

Possible syntactic structures are quantified
in Table~\ref{tab:syn} for
Possibilities on noun, verb and prepositional phrases
and in sentences.
(Counting along the number of words.)

\subsection{Synthesis of text}
Such counting exercise is also useful
for (semi-)automate writing through scripting.
The syntax organizes the words in larger structures.
The rhymes are restricted,
the ending is as a tonality in music.

There are some specific tasks for achieving texts,
such as finding the number of syllables considering the elisoes,
or handling interaction of the writer with the script
to choose sentences or verses or stanzas.

@@

Figure~\ref{fig:syn} holds some synthesized texts
with different coloring schemes for syntax highlighting,
as presented in the next section.

### 2017-11-19 article/pos.tex
adiantada escrita e scripts de analise

\begin{table*}[h!]
\begin{center}
\caption{"POS tags incident and chosen.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the text highlighting Plugin, for example,
            a token has to have an established tag to have
            a defined color.
            On the Chosen column, the tokens were regarded only once
            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}
\begin{tabular}{| l | c | c |}\hline
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\\hline
ADJECTIVE & 40  & 35 \\\hline
VERB & 18  & 13 \\\hline
PARTICLE & 12  & 12 \\\hline
PRE & 5  & 5 \\\hline
PREPOSITION & 5  & 5 \\\hline
NUMBER & 4  & 1 \\\hline
total & 142  & 120 \\\hline
\end{tabular}\end{center}
\end{table*}

>>>>

\begin{table*}[h!]
\begin{center}
\caption{"POS tags incident and chosen.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the text highlighting Plugin, for example,
            a token has to have an established tag to have
            a defined color.
            On the Chosen column, the tokens were regarded only once
            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}
\begin{tabular}{| l | c | c |}\hline
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\\hline
ADJECTIVE & 40  & 37 \\\hline
VERB & 15  & 15 \\\hline
PARTICLE & 12  & 12 \\\hline
PREPOSITION & 5  & 5 \\\hline
NUMBER & 4  & 1 \\\hline
PRE & 0  & 0 \\\hline
total & 134  & 120 \\\hline
\end{tabular}\end{center}
\end{table*}
